---
title: "Week 3. Methods for Arthropod Sampling"
date: "Sept 17 & 18, 2024"
output: 
  html_document:
    toc: TRUE
    toc_depth: 3
  pdf_document:
    toc: TRUE
    toc_depth: 3
    keep_tex: true
pkgdown:
  as_is: true
---
# Arthropod Sampling: Experimental Design

Every group will interact with each trap type (even if their group doesn’t end up using that specific data)

* 4 habitats (interior forest, edge forest, wetland, open field)
* 10 groups (5 Tues, 5 Wed)
* 5 types of traps
    * Combo traps (4 traps x 2 weeks = 8 total; 2 per habitat)
    * Pitfall traps (1 per habitat x 2 habitats per group x 10 groups = 20 total; 5 per habitat)
    * Berlese funnels (1 per habitat x 2 habitats per group x 10 groups = 20 total; 10 per habitat)
    * Sweep nets (we have 10 total, 2 available per group) – Transects? Pool samples?
    * Beat nets (we have 10 total, 2 available per group) – Transects? Pool samples?

### Week 1 field work (Sept 17-18)

* Finalize experimental design
* Visit collection site
* Collect sweep net & beat net samples in each habitat [every group]
* Deploy pitfall traps, round 1 [every group]
* Initial work on ID [every group]


### Week 2 field work (Sept 24-25)
* Deploy pitfall traps, round 2 [3 groups per lab section, each group deploys 4 traps (one per habitat)]
* Deploy combo traps, round 1 [2 groups per lab section, each group deploys 1 trap, cover all four habitats between both lab sections]
* Collect sweep net & beat net samples in each habitat [every group]
* Collect leaf litter samples & set up Berlese funnel traps [every group, every habitat = 20 samples]
* Continue ID work on previous samples

### Week 3 field work (Oct 1-2)
* Collect pitfall traps & combo traps [groups bring to central field location to swap gear]
* Deploy pitfall traps, round 3 [2 groups per lab section, each group deploys 4 traps (one per habitat)]
* Deploy combo traps, round 2 [3 groups per lab section; will need to distribute the extra group between the other two groups for 2 groups total, each group deploys 1 trap, cover all four habitats between both lab sections]
* Collect sweep net & beat net samples in each habitat [every group]
* Collect Berlese funnel samples
* Continue ID work on previous samples

### Week 4 field work (Oct 8-9)
* Collect pitfall traps & combo traps [bring back to lab]
* Collect sweep net & beat net samples in each habitat [every group]
* Clean & stow gear
* Continue ID work on previous samples 
* Data cleaning
* Data analysis
* Presentation prep

### Week 5 (Oct 15-16)
* Presentations


# Field Bag Supplies for Arthropod Sampling

### Materials Required

*	Messenger field bags
*	Field notebooks
*	Arboretum map
*	Measuring tape
*	Permanent markers
*	Pencils
*	Ziplock baggies
*	Funnel
*	One pair of clippers
*	Buckets for toting tools & traps
*	Collection containers for wet samples
*	Soapy water for setting pitfall & combo traps
    *	How to make soapy water: Add approximately 1 milliliter of dish soap to a full gallon of water. Replace jug cap and turn over to mix (avoid creating bubbles if possible).
*	Mapping app on phone
*	Level app on phone (iOS: “Measure” app; Android: Google search “Bubble level”)


# Sweep Net Sampling

### Materials Required



*	Sweep net (assembled)
*	Kill jar
    *	Mason jar
    *	Cotton ball
    *	Acetone or substitute
*	Labelling tape
*	Permanent marker

```{r sweep net, out.width='50%', out.extra='style="float:right; padding:10px"', echo = FALSE}
knitr::include_graphics(file.path(here::here(), 'man', 'figures', 'sweep_net.jpg'))
```

### Procedure:
1.	Select a location from which to sample. Your group is assigned a general area. Choose a random starting location within that parameter.
2.	Label a kill jar with the following information: Lab day, Group number, Habitat information, Trap type (=Sweep net), Date.
3.	Determine the direction of this transect. 
4.	Run your measuring tape out to at least 15 meters.
5.	You will sample five times along this transect, stopping every 3 meters to sample. Drop a GPS pin to identify each location within this transect. Share the GPS pins with your instructor.
6.	Assemble the sweep net. Add the frame/net component to the removable handle and lock the handle in place. Unfurl the net.
7.	Hold the net in your dominant hand.
8.	Place the sweep net slightly above and to the side of the shrub or tree or vegetation to be sampled 
9.	“Sweep” across the vegetation in the same direction five times. Move quickly.
10.	Collect whatever arthropods have fallen into the net by placing them in the kill jar.
11.	Repeat on the transect as instructed for this sample (instructor will share details). Pool all collected arthropods into a single kill jar for the transect.
12.	Take your sample back to the lab and transfer into storage containers as directed. Be sure to transfer label information too. This sample is ready for identification.


# Beat Net Sampling

### Materials Required

*	Beat net (assembled)
*	Beating stick (use handle from sweep net)
*	Kill jar (one for each sample to be collected)
    *	Mason jar
    *	Cotton ball
    *	Acetone or substitute 
*	Labelling tape
*	Permanent marker

```{r beat net, out.width='50%', out.extra='style="float:right; padding:10px"', echo = FALSE}
knitr::include_graphics(file.path(here::here(), 'man', 'figures', 'beat_net.jpg'))
```

### Procedure:
1.	Select a location from which to sample. Your group is assigned a general area. Choose a random starting location within that parameter.
2.	Label a kill jar with the following information: Lab day, Group number, Habitat information, Trap type (=Beat net), Date.
3.	Determine the direction of this transect. 
4.	Run your measuring tape out to at least 15 meters.
5.	You will sample five times along this transect, stopping every 3 meters to sample. Drop a GPS pin to identify each location within this transect. Share the GPS pins with your instructor.
6.	Unfold the beat net and add the wooden frame. Orient the net so that the sheet is facing up with the frame facing down.
7.	Place the beat net under the shrub or tree or vegetation to be sampled.
8.	Use the removable handle from a sweep net to “beat” the vegetation five times. Use about 50% of your strength.
9.	Collect whatever arthropods have fallen to the sheet by placing them in the kill jar.
10.	Repeat on the transect as instructed for this sample (instructor will share details). Pool all collected arthropods into a single kill jar for the transect.
11.	Take your sample back to the lab and transfer into storage containers. Be sure to transfer label information too. This sample is ready for identification.

# Berlese Funnel Sampling

### Materials required for procuring sample:

*	Ziploc baggie (one per sample)
*	Permanent marker
*	Digging tool (at least one)
    *	Trowel
    *	Shovel

```{r berlese funnel, out.width='50%', out.extra='style="float:right; padding:10px"', echo = FALSE}
knitr::include_graphics(file.path(here::here(), 'man', 'figures', 'berlese_funnel.jpg'))
```

### Procedure:
1.	Select a location from which to sample.
2.	Label a Ziploc baggie with the following information: Lab day, Group number, Habitat information, Trap type, Date.
3.	Scoop leaf litter and surface soil from your location (approximately 1-2” depth of surface soil and whatever leaf litter is on top of that, from an area approximately 6” square).
4.	Place leaf litter/soil sample into the Ziplock baggie and seal it. 
5.	Drop a GPS pin to identify the location of this trap. Share the GPS pin with your instructor.
6.	Take your sample back to the lab. This sample is ready for funneling.

### Materials required for setting up funnel:

*	Leaf litter/soil samples
*	Mason jar labeled with sample information (save lid for later use)
*	Soapy water
*	Large funnel
*	6”x6” square of wire mesh
*	6”x6” squares of cheesecloth (3-4)
*	Light source (clip-on, if possible) – placed 6” from top layer of leaf litter for 24 hours
*	Ring stand with ring

### Procedure:
1.	Set the lid of the jar aside for use later in the week. 
2.	Carefully assemble the Berlese funnel apparatus according to the image above.
3.	Gently place the sample you took in the field into the top of the funnel.
4.	Place the light source approximately 6” away from the leaf litter layer and turn it on.
5.	Label both your apparatus on the base of the ring stand and your jar of soapy water with the following information: Lab day, Group number, Habitat information, Trap type (=Berlese), Date.
6.	After 48 hours, you will collect these samples. Do so replacing the lid to the jar. Disassemble your funnel and dispose of your sample as instructed. This sample is ready for filter processing.

# Pitfall Trap Sampling

### Materials Required

For setting up this trap:

*	Pitfall trap (jar with lid)
*	Umbrella cover for trap
*	Soapy water
*	Flagging tape
*	Digging tool (at least one)
    *	Trowel
    *	Shovel
    *	Soil corer
*	Waterproof nametags w/ contact info for traps left out

```{r pitfall trap, out.width='50%', out.extra='style="float:right; padding:10px"', echo = FALSE}
knitr::include_graphics(file.path(here::here(), 'man', 'figures', 'pitfall_trap.jpg'))
```

### Procedure:
1.	Select a location from which to sample.
2.	Set the lid of the jar aside for use next week.
3.	Dig a hole deep and wide enough to fully submerge your jar such that the lip of the jar is level with the surrounding surface.
4.	Backfill around the jar so that the jar is secured and sits level. There should be no barriers to arthropods; they should be able to crawl up (and into) the trap. Use your phone’s leveling app if necessary. Be careful not to let dirt fall into the jar, as this will complicate processing of your sample later on.
5.	Disguise the trap by adding some leaf litter around it so that it appears to match the neighboring ground.
6.	Gently place the umbrella cover directly above the jar to act as a barrier for falling debris and rain. Secure the umbrella cover by pressing it into the ground a few inches.
7.	Add flagging tape on a nearby shrub or tree to help you find the trap again one week from now.
8.	Add the contact information card to the flagging tape to notify passersby not to disturb the trap.
9.	Drop a GPS pin to identify the location of this trap. Share the GPS pin with your instructor.
10.	After one week, you will collect these samples. Do so by carefully removing the jar from the ground, replacing the lid, and labelling it with the following information: Lab day, Group number, Habitat information, Trap type (=Pitfall trap), Date. Fill the hole in as best you can and remove the flagging tape.
11.	Take your sample back to the lab. This sample is ready for filter processing.

# Malaise Combo Flight Intercept Trap Sampling

### Materials Required

Parts of Combo Trap

*	Mini Bungee cords for keeping PVC parts together during transport (3-4 per trap) 
*	Top trap: Modified storage container
*	Bottom trap: Aluminum roaster pan spraypainted yellow
*	Nose cone apparatus
    *	Funnel portion of 2-liter soda bottle (connects cone net to top trap)
    *	54” square piece of white tulle netting
    *	Binder clips to anchor netting
    *	Field flags (wire frame for supporting nose cone)
    *	Rubber bands (stabilizes netting)
*	Intercept net & frame
    *	Four pieces of 3/4” PVC and 
    *	Four 3-prong PVC connectors
    *	Intercept net (white tulle netting reinforced with hem tape)
*	4 pieces of pink rebar (stabilizes intercept net frame)
*	Rope stability lines: 4 lengths, each 8’-10 each
*	4 tent stakes: One per stability line
*	Soapy water
*	Flagging tape
*	Hammer or mallet
*	Containers for collecting combo trap samples to carry back to lab: Separate collection jars for top vs bottom traps
*	Waterproof nametags w/ contact info for traps left out


```{r malaise combo trap, out.width='50%', out.extra='style="float:right; padding:10px"', echo = FALSE}
knitr::include_graphics(file.path(here::here(), 'man', 'figures', 'malaise_combo_trap.jpg'))
```

### Procedure:
1.	Select a location from which to sample.
2.	Assemble the combo trap. Your instructor or TA will be available to assist. 
3.	Once complete, the trap must be as level as possible to stabilize the top trap. Use your phone’s leveling app if necessary. 
4.	Carefully add soapy water to the top (1” depth) and bottom (2” depth) traps. Be careful not to let dirt or debris fall into the traps, as this will complicate processing of your sample later on. Adjust the trap for stability as necessary.
5.	Add flagging tape on a nearby shrub or tree to help you find the trap again one week from now.
6.	Add the contact information card to the flagging tape to notify passersby not to disturb the trap.
7.	Drop a GPS pin to identify the location of this trap. Share the GPS pin with your instructor.
8.	After one week, you will collect these samples. 
a.	Bottom trap: Carefully pour the soapy water/arthropod mixture into a collecting jar. Label the jar with the following information: Lab day, Group number, Habitat information, Trap type (=Combo trap), Trap component (=Top trap or Bottom trap), Date. 
b.	Top trap: Do so by carefully removing the jar from the ground, replacing the lid, and labelling it with the following information: Lab day, Group number, Habitat information, Trap type (=Combo trap), Trap component (=Top trap or Bottom trap), Date. 
9.	Remove the flagging tape from the area and carefully disassemble the trap.
10.	Take your sample back to the lab. This sample is ready for filter processing.

# Processing & Identification of Arthropod Samples

### Materials Required

#### For filter processing samples:
*	Sample will be preserved in soapy water in Mason jar (with lid) or other container
*	Coffee filter
*	Large funnel
*	Ring stand with ring
*	Glass stirring rod(s)
*	Water in squirt bottles to wet filters/speed filtering
*	Note: Pitfall trap and Berlese funnel samples can be slow to filter if contaminated with dirt; carefully use glass rod to break up clumps

```{r sample filtration, out.width='40%', out.extra='style="float:right; padding:10px"', echo = FALSE}
knitr::include_graphics(file.path(here::here(), 'man', 'figures', 'sample_filtration.jpg'))
```

#### Filtering procedure:
1.	Carefully assemble the filtering apparatus according to the image at right.
2.	Gently pour your soapy water/arthropod sample into the funnel. The goal here is to extract your arthropod specimens and remove the solution.
3.	Discard the filtered solution (may be done partway through your filtering process depending on how much solution you have).
4.	Transfer your arthropod samples to temporary sorting trays labelled with the following information: Lab day, Group number, Trap type, Habitat information, Date. Avoid transferring dirt and debris if possible. You may wish to let your samples dry before transferring them.

#### For identification process:
*	Ethanol in squirt bottles to break up dirt clumps/clean specimens
*	Stereoscope
*	Stereoscope camera attachment
*	Tools for manipulating specimens (probe, ultra-fine forceps, ultra-fine paintbrushes, cardstock or index cards)
*	Labelling tape
*	Sorting trays (4)
*	Storage box
*	ID books and other sources (websites, etc.)


#### Identification procedure:
1.	Work on one trap at a time.
2.	Sort arthropods into morphospecies.
3.	ID each morphospecies to Order and Family.
4.	Record number of individuals in each Order/Family group.


# Collaborative Data Set

The entire class is contributing data toward one large, collaborative data set. Your group will have the freedom to use subsets of that data to answer your own questions.

Some of the data in this set are categorized as explanatory variables (E) and others are response variables (R). Which are which?

Categories in this data set:

| Lab Day’ |	Group Number” |	Habitat Type* |	Trap Type** |	Date | Trap | Set |	Date Trap Collected |	Order |	Family |	Number of Individuals |
| :---: |	:---: |	:---: |	:---: |	:---: |	:---: |	:---: | :---: |	:---: |
| ? |	? |	? |	? |	? |	? |	? | ? |	? |
|  |	 |	 |	 |	 |	 |	 | |	 |

Options:
’ Tues or Wed
” 1, 2, 3, 4, or 5
* Forest edge, forest interior, riparian, open field
** Sweep net, beat net, pitfall trap, Berlese funnel, combo trap




